### 2025-02-03 paper/preprint/paper.tex
intro basically done

Part of the problem is that great degrees of freedom provide room for
many locally optimal solutions when using narrow objectives
(\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on predictive
performance, these solutions may seem to provide compelling explanations
for the data at hand, even though they are in fact not grounded in
meaningful semantics. In other words, the greatest strength of modern
representation learning is also its greatest pitfall.

This opaqueness has another dire consquence: since we cannot easily
interpret the mapping from inputs to outputs, deploying such models in
practive effectively means blindly relying on model predictions. On
countless occasions, this has already caused real harm to people who
were affected adversely and often unfairly by automated decision-making
systems involving opaque models (\citeproc{ref-oneil2016weapons}{O'Neil
2016}). Prominent voices have therefore argued to completely abolish the
use of such ``black boxes'' along with any attempts to explain their
predictions (\citeproc{ref-rudin2019stop}{Rudin 2019}): after all, if we
had some interpretable abstraction of a ``black box'' that explains its
decisions with full fidelity, there would be no need for the ``black
box'' at all, because we could simply rely on the abstraction for
decision-making.

While we sympathise with that stance and share the aforementioned
concerns, the reality is that the trend towards complexity has not been
reversed. If anything, it has accelerated on the back of unprecented
investments in computing infrastructure. With opaque AI here to stay, we
believe that \textbf{explainable AI} (XAI) will conintue to play an
important role in dealing with ``black boxes''.

@@

Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
counterfactual images for MNIST and ImageNet through independent
mechanisms (IM): each IM learns class-conditional input distributions
over a specific lower-dimensional, semantically meaningful factor, such
as \emph{texture}, \emph{shape} and \emph{background}. The demonstrate
that using these generated counterfactuals during classifier training
improves model robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization.

@@

\bibitem[\citeproctext]{ref-rudin2019stop}
Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
Models for High Stakes Decisions and Use Interpretable Models
Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
\url{https://doi.org/10.1038/s42256-019-0048-x}.

@@

>>>>

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on
predictive performance, these solutions may seem to provide compelling
explanations for the data, when in fact they are based on purely
associative, semantically meaningless patterns. This poses two related
challenges: firstly, it makes these models inherently opaque, since
humans cannot simply interpret what type of explanation the complex
learned representations correspond to; secondly, even if we could
resolve the first challenge, it is not obvious how to mitigate models
from learning representations that correspond to meaningless and
undesirable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems involving opaque
models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
tools can aide us in monitoring models and providing affected
individuals with recourse
(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
Russell 2017}).

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are deemed desirable by humans. Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both desirable and valid for any given
model, independent of whether the learned representations are also
compatible with undesirable explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

In this work, we build on these findings and introduce a
\textbf{counterfactual training}: a novel approach geared explicitly
towards aligning model representations with desirable explanations.

From this perspective, adversarial training induces models to
``unlearn'' representations that are susceptible to the semantically
most meaningless explanations---adversarial examples.

@@

Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
counterfactual images for MNIST and ImageNet through independent
mechanisms (IM): each IM learns class-conditional input distributions
over a specific lower-dimensional, semantically meaningful factor, such
as \emph{texture}, \emph{shape} and \emph{background}. They demonstrate
that using these generated counterfactuals during classifier training
improves model robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization.

@@

@@

\bibitem[\citeproctext]{ref-schut2021generating}
Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.

### 2025-02-03 paper/sections/intro.qmd
intro basically done

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so they have also learned increasingly complex representations that we can no longer easily interpret.

This trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are highly sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts. Despite continued research efforts towards making models more robust to such counterfactuals, the issue remains unresolved even for models that are considered to small by today's standards [@kolter2023keynote]. It seems that robust representation learning is fundamentally at odds with large degrees of freedom.

@wilson2020case

@rudin2019stop

>>>>

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

This trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts. Despite partially effective mitigation strategies such as **adversarial training**, truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]. Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and undesirable explanations.

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons]. Effective XAI tools can aide us in monitoring models and providing affected individuals with recourse [@wachter2017counterfactual].

To our surprise, the second challenge has not yet attracted any consolidated research effort. Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are deemed desirable by humans. Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].

Building on these findings, we introduce a **counterfactual training**: a novel approach geared explicitly towards aligning model representations with desirable explanations.

From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.
